---
title: "Clip-SVT versus conditional logistic collaborative filtering"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  config: "quick"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5, dpi = 110)
# `params` only exists under rmarkdown::render. Defining a fallback lets the
# same file be knitted to markdown with knitr::knit when pandoc is unavailable,
# which is the case on machines without RStudio.
if (!exists("params")) {
  params <- list(config = Sys.getenv("EXP_CONFIG", "quick"))
}
library(ggplot2)
library(dplyr)
theme_set(theme_bw(base_size = 11))

res <- function(stem) {
  f <- file.path("results", sprintf("%s_%s.rds", stem, params$config))
  if (!file.exists(f)) return(NULL)
  readRDS(f)
}

method_labs <- c(baseline = "intercept only", old = "old (conditional LCF)",
                 new_cp = "new (Clip-SVT, Cp)",
                 new_boot = "new (Clip-SVT, Brier+cov)",
                 new_fixed = "new (Clip-SVT, fixed alpha)",
                 new_tuned = "new (Clip-SVT, tuned alpha)",
                 new = "new (Clip-SVT)")
pal <- c("intercept only" = "grey55", "old (conditional LCF)" = "#C2453E",
         "new (Clip-SVT, Cp)" = "#3A6FB0",
         "new (Clip-SVT, Brier+cov)" = "#2E8B72",
         "new (Clip-SVT, fixed alpha)" = "#3A6FB0",
         "new (Clip-SVT, tuned alpha)" = "#2E8B72",
         "new (Clip-SVT)" = "#3A6FB0")
lab <- function(x) factor(method_labs[x], levels = names(pal))
```

Config: **`r params$config`**.

Both methods are scored on the same estimand, the marginal survival
$S^{(t)} = P(\text{value} > d_t)$. The old method estimates conditional
probabilities and multiplies them up; the new method estimates
$\sigma(X^{(t)})$ directly. The intercept-only fit is shown throughout as the
floor that any factorization has to beat.

# 1. Accuracy

```{r accuracy-load}
acc <- res("accuracy")
```

```{r accuracy-overall, eval = !is.null(acc)}
overall <- acc$summary %>%
  filter(is.na(t)) %>%
  mutate(method = lab(fit_name))

overall %>%
  group_by(method, shape, pi_tail) %>%
  summarise(rmse = mean(rmse), cross_entropy = mean(cross_entropy),
            auc = mean(auc, na.rm = TRUE), secs = mean(elapsed),
            .groups = "drop") %>%
  arrange(shape, pi_tail, rmse)
```

## Error against the truth, by tail sparsity

The design axis that isolates the mechanism. Lower `pi_tail` means the last
mask is sparser, so the old method's conditional risk set is more starved.

```{r accuracy-tail, eval = !is.null(acc)}
ggplot(overall, aes(factor(pi_tail), rmse, fill = method)) +
  geom_boxplot(outlier.size = 0.6, position = position_dodge(0.8)) +
  facet_wrap(~ shape, scales = "free_y") +
  scale_fill_manual(values = pal) +
  labs(x = "prevalence of the last mask", y = "RMSE of estimated survival",
       fill = NULL) +
  theme(legend.position = "bottom")
```

## Error by threshold

```{r accuracy-by-t, eval = !is.null(acc)}
acc$summary %>%
  filter(!is.na(t)) %>%
  mutate(method = lab(fit_name)) %>%
  group_by(method, t, Tn, pi_tail) %>%
  summarise(rmse = mean(rmse), .groups = "drop") %>%
  ggplot(aes(t, rmse, colour = method)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  facet_grid(pi_tail ~ Tn, scales = "free", labeller = label_both) +
  scale_colour_manual(values = pal) +
  labs(x = "threshold index t", y = "RMSE", colour = NULL) +
  theme(legend.position = "bottom")
```

The engines tie at `t = 1`, where they see identical information, and separate
as the old method's conditioning compounds across thresholds.

## Calibration and monotonicity

```{r accuracy-cal, eval = !is.null(acc)}
acc$summary %>%
  filter(!is.na(t)) %>%
  mutate(method = lab(fit_name)) %>%
  group_by(method) %>%
  summarise(`calibration slope` = mean(cal_slope, na.rm = TRUE),
            `calibration intercept` = mean(cal_intercept, na.rm = TRUE),
            `monotonicity violations` = mean(mono_violation),
            .groups = "drop")
```

## Error by threshold: new versus old

The per binary mask comparison, scored as squared error against the **true**
survival probabilities. Selection no longer holds entries out, so the Brier
score against the observed `Y` would now be in sample -- and the new engine fits
`Y` better in sample by construction, because it retains more rank. Scoring
against the truth needs no split and is immune to that.

```{r accuracy-brier-by-t, eval = !is.null(acc)}
acc$summary %>%
  filter(!is.na(t)) %>%
  mutate(method = lab(fit_name)) %>%
  group_by(method, t, shape) %>%
  summarise(sq_err = mean(rmse^2), .groups = "drop") %>%
  ggplot(aes(t, sq_err, colour = method)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  facet_wrap(~ shape, scales = "free_y") +
  scale_colour_manual(values = pal) +
  labs(x = "threshold index t",
       y = expression(mean((hat(pi) - S[true])^2)), colour = NULL) +
  theme(legend.position = "bottom")
```

```{r accuracy-brier-tab, eval = !is.null(acc)}
acc$summary %>%
  filter(!is.na(t)) %>%
  mutate(method = lab(fit_name)) %>%
  group_by(method, t) %>%
  summarise(sq_err = mean(rmse^2), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = method, values_from = sq_err)
```

## Paired comparison against the old engine

Within a `rep` the seed is a function of the design, so every engine sees the
same simulated data. Differences are therefore **paired**, which is both the
right test and far more powerful than comparing two marginal means. Negative
`d_rmse` means the new method is more accurate; negative `d_secs` means faster.

**Read `p_t`, not `p_wilcox`.** With `n` replicates the signed rank test's
smallest attainable two sided p is `2 / 2^n` -- 0.0625 at `n = 5`. A `p_wilcox`
of 0.0625 therefore means *every* pair favoured one side and the test has
saturated; it is the strongest evidence that test can give at this `n`, not weak
evidence. `wins_rmse` out of `n` is the honest non-parametric summary.

```{r accuracy-paired, eval = !is.null(acc)}
wide <- acc$summary %>%
  filter(is.na(t)) %>%
  select(shape, rep, fit_name, rmse, elapsed) %>%
  tidyr::pivot_wider(names_from = fit_name, values_from = c(rmse, elapsed))

paired <- function(new) {
  d_r <- wide[[paste0("rmse_", new)]] - wide$rmse_old
  d_s <- wide[[paste0("elapsed_", new)]] - wide$elapsed_old
  data.frame(
    arm = new, shape = wide$shape, d_rmse = d_r, d_secs = d_s)
}
# `new_boot` is absent whenever cfg$selection$do_boot is FALSE, which is the
# default now, so take whichever tuned arms the run actually produced.
arms <- intersect(c("new_cp", "new_boot"),
                  sub("^rmse_", "", grep("^rmse_", names(wide), value = TRUE)))
pd <- do.call(rbind, lapply(arms, paired))

# NOTE: summarise() evaluates its arguments in order and each one sees the
# results of the previous ones. Naming an output `d_rmse` while later arguments
# still need the *column* `d_rmse` silently replaces the vector with its own
# mean, so sd() returns NA and sum(< 0) returns 1. Every summary here therefore
# gets a name distinct from the columns it reads.
pd %>%
  group_by(arm, shape) %>%
  summarise(n = dplyr::n(),
            mean_d_rmse = mean(d_rmse), se_rmse = sd(d_rmse) / sqrt(dplyr::n()),
            wins_rmse = sum(d_rmse < 0),
            mean_d_secs = mean(d_secs), se_secs = sd(d_secs) / sqrt(dplyr::n()),
            wins_secs = sum(d_secs < 0),
            p_wilcox = tryCatch(
              stats::wilcox.test(d_rmse)$p.value, error = function(e) NA_real_),
            p_t = tryCatch(
              stats::t.test(d_rmse)$p.value, error = function(e) NA_real_),
            .groups = "drop") %>%
  as.data.frame()
```

# 1b. Choosing alpha

Two criteria, both scored against an `oracle`: the grid alpha minimising RMSE
against the true survival probabilities. That reference exists only in
simulation, which is why this lives here. The held out rule that used to be
stage 1 is gone -- selection no longer splits entries.

`cp_note_asis` and `cp_hard_df` are controls, not candidates. They are the
Mallows Cp surrogate without the two corrections it needs -- a variance
re-estimated at every alpha, and a hard parameter count that ignores how much
soft thresholding shrank what it retained.

```{r selection-load}
sel <- acc$selection
selp <- acc$sel_path
selr <- acc$sel_ranks
stab <- acc$sel_stability
```

```{r selection-table, eval = !is.null(sel)}
sel %>%
  group_by(stage) %>%
  summarise(`signed k - oracle` = mean(k_minus_oracle),
            `|k - oracle|` = mean(abs_k_minus_oracle),
            `exact hits %` = 100 * mean(k_minus_oracle == 0),
            rmse = mean(rmse_all),
            `excess rmse` = mean(excess_rmse),
            secs = mean(elapsed),
            .groups = "drop") %>%
  arrange(`|k - oracle|`)
```

## Why the surrogate needs its two corrections

`v_tilde` is the mean Bernoulli variance of the fitted probabilities. As the
shrinkage weakens the fit drives them toward 0 and 1, so a variance re-estimated
per alpha collapses exactly where the degrees of freedom explode, and the
penalty it multiplies vanishes with it.

```{r selection-vtilde, eval = !is.null(selp)}
selp %>%
  group_by(k, alpha) %>%
  summarise(v_tilde = mean(v_tilde), df = mean(df), df_hard = mean(df_hard),
            .groups = "drop") %>%
  tidyr::pivot_longer(c(v_tilde, df, df_hard)) %>%
  ggplot(aes(alpha, value)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  scale_x_log10() +
  facet_wrap(~ name, scales = "free_y") +
  labs(x = "alpha", y = NULL)
```

## Rank recovery, per threshold

The direct answer to whether the right rank is being chosen. Note the caveat:
soft thresholding biases retained singular values downward, so the alpha that
minimises estimation error keeps *more* components than the truth. A stage can
recover the rank better while estimating worse, which is why the table above is
reported beside this.

```{r selection-ranks, eval = !is.null(selr)}
selr %>%
  group_by(stage, t) %>%
  summarise(fitted = mean(rank_fit), truth = mean(rank_true), .groups = "drop") %>%
  ggplot(aes(t, fitted, colour = stage)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  geom_line(aes(y = truth), colour = "black", linetype = 2, linewidth = 0.6) +
  labs(x = "threshold index t", y = "retained rank (dashed = truth)",
       colour = NULL) +
  theme(legend.position = "bottom")
```

```{r selection-rank-tab, eval = !is.null(sel)}
sel %>%
  group_by(stage) %>%
  summarise(`mean fitted rank` = mean(mean_rank),
            `mean true rank` = mean(mean_true_rank),
            `mean signed error` = mean(rank_err),
            `mean |error|` = mean(abs_rank_err), .groups = "drop") %>%
  arrange(`mean |error|`)
```

## The criterion along the path

```{r selection-path, eval = !is.null(selp)}
selp %>%
  group_by(k, alpha) %>%
  summarise(rmse_all = mean(rmse_all), .groups = "drop") %>%
  ggplot(aes(alpha, rmse_all)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  scale_x_log10() +
  labs(x = "alpha", y = "RMSE against the truth, all entries")
```

## Does the bootstrap answer depend on where it was anchored?

The resample is drawn from the fit at the anchor, so an over-smoothed anchor
carries more Bernoulli variance than the truth does. Every candidate then chases
more noise than it really would, the weakly shrunk ones most of all. If that
matters, moving the anchor moves the winner.

```{r selection-stability, eval = !is.null(stab)}
stab %>%
  group_by(anchor, k) %>%
  summarise(M = mean(M), .groups = "drop") %>%
  ggplot(aes(k, M, colour = factor(anchor))) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  labs(x = "candidate k", y = "M(alpha)", colour = "anchor k")
```

```{r selection-stability-tab, eval = !is.null(stab)}
stab %>%
  distinct(shape, Tn, pi_tail, rep, anchor, argmin) %>%
  count(anchor, argmin, name = "cells")
```

## Arithmetic versus parallel scaling

Stage 01 times both engines at the same `ncores`, which is the wall clock you
would see. That conflates two things: how much work an algorithm does, and how
well that work spreads across cores. The two differ by construction -- the
bootstrap has `B` independent replicates, `cv.logisticcfR` only
`length(lambdas)` fits per rank level -- so the same table at `ncores = 1`
separates them. `speedup` is the serial time over the parallel time.

Two things to read carefully here.

**`new_cp` does not use `ncores` at all.** `ncores` reaches only
`boot_covariance()`; `bmfsvt_path()` is serial. Its `speedup` column is
therefore pure run to run noise, and it is large (about 1.4x), which sets the
precision of every single-shot timing on this page. The five-replicate means in
stage 01 are the numbers to quote, not these.

**The old engine gains nothing from cores**, which is what makes stage 01's
equal-`ncores` comparison fair rather than generous. Spreading three lambdas per
rank level does not repay the fork, and on the square shape it is *slower* with
eight cores than with one.

```{r serial-timing, eval = !is.null(res("serial_timing"))}
st <- res("serial_timing")
st %>%
  tidyr::pivot_longer(dplyr::any_of(c("old", "new_cp", "new_boot")),
                      names_to = "arm", values_to = "secs") %>%
  tidyr::pivot_wider(names_from = ncores, values_from = secs,
                     names_prefix = "cores_") %>%
  mutate(speedup = .data[[grep("^cores_", names(.), value = TRUE)[1]]] /
                   .data[[grep("^cores_", names(.), value = TRUE)[2]]]) %>%
  arrange(shape, arm) %>%
  as.data.frame()
```

---

## Retired stages

`02_timing.R`, `03_constraint.R` and `05_merge.R` were removed; the suite is now
`01_accuracy.R` -> `04_render.R`, with `06_serial_timing.R` as an optional
check. Both retired stages had reached a settled answer, recorded here so the
result outlives the code (recoverable with `git log --diff-filter=D`):

- **The clipping heuristic is sufficient; ADMM is not needed.** Against an exact
  CVXR solve of the constrained program the two agreed to within an optimization
  error far smaller than the statistical error the method makes anyway. `b` and
  `d` from Algorithm 2 measure constraint *activity*, not heuristic failure --
  at a constrained optimum the unconstrained gradient step must leave the
  feasible set, so they take nearly the same values at an exact solution. Do not
  threshold them to judge the clipping; compare against an exact solve.
- **An RcppArmadillo port of the elementwise kernels is worth doing.** Profiled
  at 800 x 200 with `T = 10` *after* the `soft_svt()` rework: SVD 40%,
  elementwise kernels 47%, other 13%. Before that rework the split was 65% / 27%
  and argued the other way, so re-profile after any change to the decomposition
  path before revisiting.
